Different rules for binocular combination of luminance flicker in cortical and subcortical pathways

### 2023-06-12 manuscript.Rmd
Started implementing reviewer suggestions

processdata <- 2  # this flag determines the amount of processing, with 4 levels:
# 0 - do no processing, generate the pdf using existing versions of all figures
# 1 - generate figures using the processed group data
# 2 - average data across participants, do bootstrapping
# 3 - download all raw data and analyse (requires 48GB of storage)

@@

The brain must combine information across multiple sensory inputs to derive a coherent percept of the external world. This involves a process of signal combination both within [@Baker2017] and between [@Ernst2002] the senses. Binocular vision is a useful test case for signal combination, as the inputs to the two eyes overlap substantially (in species with forward-facing eyes), and the neural locus is well-established [@Hubel1962]. Much of our knowledge about binocular combination derives from studies on the contrast response of the 'canonical' visual pathway, in which signals pass from the eyes to primary visual cortex (V1), via the lateral geniculate nucleus (LGN) [@Purves2008]. However, signals are also combined across the eyes in the network of subcortical nuclei that govern pupil diameter in response to absolute light levels [@McDougal2008] and much less is known about the computations that operate in these subcortical pathways. In addition, our primary purpose here is to investigate the computations governing signal combination in these two anatomically distinct pathways in response to luminance changes.

For pattern vision, binocular presentation confers greater sensitivity to low contrast targets than monocular presentation. This is known as binocular summation, with summation ratios (the relative improvement under binocular presentation) at detection threshold lying between $\sqrt{2}$ and 2 [@Campbell1965; @Baker2018]. This advantage is lost at high stimulus intensities, where both psychophysical performance (contrast discrimination thresholds) [@Legge1984; @Meese2006] and neural activity [@Moradi2009; @Baker2017] are approximately equal for monocular and binocular presentation. Contemporary models of binocular vision [@Ding2006; @Meese2006] advocate a process of interocular suppression that normalizes the two eyes inputs at high contrasts and negates the binocular advantage. This is consistent with our everyday experience of 'ocularity invariance' [@Baker2007]: perceived contrast does not change when one eye is opened and closed.

@@

The stimuli were two discs of achromatic flickering light with a diameter of 3.74 degrees, presented on a black background. The same stimuli were used for all three experiments. Four dark red lines were added around both discs to help with their fusion into one binocular disc (see insert in Figure \@ref(fig:pupildata)b for an example of the fused stimulus). The discs were viewed through a four-mirror stereoscope, which used front silvered mirrors to avoid internal reflections, and meant that participants saw a single fused disc. The use of a stereoscope allowed us to modulate the stimuli in three different ocular configurations: monocular, binocular, and dichoptic. Note that during monocular presentation of flicker, the unstimulated eye still saw the static (non-flickering) disc of mean luminance.

@@

Before each experiment, participants calibrated the stereoscope by adjusting the angle of the mirrors. This was done so that they would perceive the two discs as one fused disc when looking at the screen through the stereoscope.

@@

We included six distinct ocular conditions, each at five temporal contrast levels (combined factorially) relative to the mean luminance: 6, 12, 24, 48 and 96\%. Contrast was defined as temporal Michelson contrast; the difference between maximum and minimum luminances, scaled by the mean and expressed as a percentage. In the first three conditions, the discs flickered at 2 Hz, in either a monocular, binocular, or dichoptic arrangement. In the dichoptic condition the non-target eye saw a fixed contrast of 48\%. In the remaining three conditions (termed the cross-frequency conditions) one eye's disc flickered at 1.6Hz, and the other eye's disc flickered at 2Hz. We also tested monocular (one eye sees 1.6Hz flicker, the other sees mean luminance), binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 2Hz, mask contrast of 48\% at 1.6Hz in the other eye) arrangements. The rationale for flickering both eyes at 2Hz is that we can then measure summation behaviour between the eyes in the pupil and EEG response at 2Hz. The rationale for flickering the eyes at different frequencies is that this permits measurement of suppression between the eyes (i.e. the reduction in the 2Hz response when a 1.6Hz mask component is added to the other eye). We counterbalanced presentation of the target stimulus across the left and right eyes.

@@

The experiment was conducted in a darkened room with a blacked-out window. The display equipment (monitor and stereoscope) were the same as for the two previous experiments, but no EEG or pupillometry data were collected. A two-interval contrast matching procedure was used to collect data. In one interval, participants were presented with a standard fused disc that flickered at a set contrast level (either 24 or 48\%), which was selected by the experimenter at the beginning of each block. In the other interval, a target disc flickering at varying contrast levels was displayed. The contrast level of the target was controlled by a 1-up, 1-down staircase moving in logarithmic (dB) steps of contrast. The ratio of flicker amplitudes in the left and right eyes was varied across blocks and was set to be 0, 0.25, 0.5, 0.75 or 1 (9 distinct conditions). The standard and target discs were displayed for 1 second each, with an interstimulus interval of 0.5 seconds. After both discs had appeared on screen, the participants had to indicate which interval they perceived as having the more intense flicker. The intervals were randomly ordered, and all discs flickered at a frequency of 2 Hz (two cycles in sine phase).

@@

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software [@Kassner2014], which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit a single sine-wave at the fundamental frequency, rather than using Fourier analysis [e.g. @Spitschan2014]. The Fourier approach is more robust to noise at other frequencies and has been used in some previous studies [see @Barrionuevo2014; @Barrionuevo2016]. Additionally, it makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis [e.g. @Figueira2022].

@@

save(file=paste0(datadir,'Averagedata.RData'),list=c('cleanmeansP','cleanmeansPCI','cleanmasksP','cleanmasksPCI','cleanmeansEav','cleanmeansEavCI','cleanmasksEav','cleanmasksEavCI','cleanmeansE2av','cleanmeansE2avCI','cleanmasksE2av','cleanmasksE2avCI','cleanmeansE','cleanmeansECI','cleanmasksECI','cleanmeansE2','cleanmeansE2CI','cleanmasksE2','cleanmasksE2CI','meanspectraP','meanspectraPCI','meanspectraE','meanspectraECI','meanwavesP','meanwavesE','meanwavesPCI','meanwavesECI','electrodes','allmeansP','allmeansE','allmeansE2'))

@@

@@

```{r pupildata, fig.cap="Summary of pupillometry results for N=30 participants. Panel (a) shows a group average waveform for binocular presentation (low pass filtered at 5Hz), with the driving signal plotted at the foot. Panel (b) shows the average Fourier spectrum, with an inset image illustrating the stimulus appearance (upper right). Panels (c,d) show contrast response functions at 2Hz for different conditions. Panel (e) shows contrast response functions at 1.6Hz for three conditions. Shaded regions and error bars indicate bootstrapped standard errors.", fig.align="center", echo=FALSE}

@@

Figure \@ref(fig:pupildata)e shows responses at 1.6Hz, for the same conditions, as well as for a condition in which a monocular stimulus flickered at 1.6Hz (grey circles). Surprisingly there again appears to be a slight facilitation effect in the binocular cross condition, particularly at lower contrasts. The dichoptic cross condition does not show clear modulation with target contrast.

@@

```{r BSratios, fig.cap="Ratio of binocular to monocular response for three data types. Each ratio is the average of ratios for N=30 participants, and error bars indicate bootstrapped standard errors.", fig.align="center", out.width = '50%', echo=FALSE}

@@

```{r TFdata, fig.cap="Binocular facilitation at different temporal frequencies. Panel (a) shows Fourier spectra for responses to binocular flicker at 5 different frequencies (offset vertically for clarity). Panel (b) shows the response at each stimulation frequency for monocular (red circles) and binocular (blue squares) presentation. Panel (c) shows the ratio of binocular to monocular responses. Error bars and shaded regions indicate bootstrapped standard errors across N=12 participants.", fig.align="center", echo=FALSE}

@@

```{r matchingdata, fig.cap="Contrast matching functions. Dotted and dashed lines are predictions of canonical summation models with a linear exponent (dotted) or an infinite exponent (dashed). Error bars indicate the standard error across participants (N=10), and are constrained along radial lines converging at the origin. Note that, for the {48\\%} match, the data point on the x axis falls higher than {100\\%} contrast. This is because the psychometric function fits for some individuals were interpolated such that the PSE fell above {100\\%}, shifting the mean slightly above that value.", out.width = '50%', fig.align="center", echo=FALSE}

@@

Pupil size determines the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2% of overall diameter [e.g. @Spitschan2014]. It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms [e.g. @Bradley2008]. We think it unlikely that these temporally coarse processes would have a differential effect on e.g. monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the two pathways are effectively distinct, but hope to investigate this directly in future neuroimaging work.

>>>>

processdata <- 0  # this flag determines the amount of processing, with 4 levels:
# 0 - do no processing, generate the pdf using existing versions of all figures
# 1 - generate figures using the processed group data
# 2 - average data across participants, do bootstrapping
# 3 - download all raw data and analyse (requires 48GB of storage)

@@

The brain must combine information across multiple sensory inputs to derive a coherent percept of the external world. This involves a process of signal combination both within [@Baker2017] and between [@Ernst2002] the senses. Binocular vision is a useful test case for signal combination, as the inputs to the two eyes overlap substantially (in species with forward-facing eyes), and the neural locus is well-established [@Hubel1962]. Much of our knowledge about binocular combination derives from studies on the contrast response of the 'canonical' visual pathway, in which signals pass from the eyes to primary visual cortex (V1), via the lateral geniculate nucleus (LGN) [@Purves2008]. However, signals are also combined across the eyes in the network of subcortical nuclei that govern pupil diameter in response to absolute light levels [@McDougal2008] and much less is known about the computations that operate in these subcortical pathways. Our primary purpose here is to investigate the computations governing signal combination in these two anatomically distinct pathways in response to luminance changes.

For pattern vision, binocular presentation confers greater sensitivity to low contrast targets than monocular presentation. This is known as binocular summation, with summation ratios (the relative improvement under binocular presentation) at detection threshold lying between $\sqrt{2}$ and 2 [@Campbell1965; @Baker2018]. This advantage is lost at high stimulus contrasts, where both psychophysical performance (contrast discrimination thresholds) [@Legge1984; @Meese2006] and neural activity [@Moradi2009; @Baker2017] are approximately equal for monocular and binocular presentation. Contemporary models of binocular vision [@Ding2006; @Meese2006] advocate a process of interocular suppression that normalizes the two eyes' inputs at high contrasts and negates the binocular advantage. This is consistent with our everyday experience of 'ocularity invariance' [@Baker2007]: perceived contrast does not change when one eye is opened and closed.

@@

The stimuli were two discs of achromatic flickering light with a diameter of 3.74 degrees, presented on a black background. The same stimuli were used for all three experiments. Four dark red lines were added around both discs to help with their perceptual fusion, giving the appearance of a single binocular disc (see insert in Figure \@ref(fig:pupildata)b for an example of the fused stimulus). The discs were viewed through a four-mirror stereoscope, which used front silvered mirrors to avoid internal reflections, and meant that participants saw a single fused disc. The use of a stereoscope allowed us to modulate the stimuli in three different ocular configurations: monocular, binocular, and dichoptic. Note that during monocular presentation of flicker, the unstimulated eye still saw the static (non-flickering) disc of mean luminance.

@@

Before each experiment, participants adjusted the angle of the stereoscope mirrors to achieve binocular fusion. This was done so that they would perceive the two discs as one fused disc when looking at the screen through the stereoscope.

@@

We included six distinct ocular conditions, each at five temporal contrast levels (combined factorially) relative to the mean luminance: 6, 12, 24, 48 and 96\%. Contrast was defined as temporal Michelson contrast; the difference between maximum and minimum luminances, scaled by the mean and expressed as a percentage. In the first three conditions, the discs flickered at 2 Hz, in either a monocular, binocular, or dichoptic arrangement. In the dichoptic condition the non-target eye saw a fixed temporal contrast of 48\%. In the remaining three conditions (termed the cross-frequency conditions) one eye's disc flickered at 1.6Hz, and the other eye's disc flickered at 2Hz. We also tested monocular (one eye sees 1.6Hz flicker, the other sees mean luminance), binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 2Hz, mask contrast of 48\% at 1.6Hz in the other eye) arrangements. The rationale for flickering both eyes at 2Hz is that we can then measure summation behaviour between the eyes in the pupil and EEG response at 2Hz. The rationale for flickering the eyes at different frequencies is that this permits measurement of suppression between the eyes (i.e. the reduction in the 2Hz response when a 1.6Hz mask component is added to the other eye). We counterbalanced presentation of the target stimulus across the left and right eyes.

@@

The experiment was conducted in a darkened room with a blacked-out window. The display equipment (monitor and stereoscope) were the same as for the two previous experiments, but no EEG or pupillometry data were collected. A two-interval contrast matching procedure was used to collect data. In one interval, participants were presented with a standard fused disc that flickered at a set contrast level (either 24 or 48\%), which was selected by the experimenter at the beginning of each block. In the other interval, a target disc was displayed, flickering at different contrast levels on each trial, but with a fixed interocular contrast ratio across the block. The contrast level of the target was controlled by a 1-up, 1-down staircase moving in logarithmic (dB) steps of contrast. The ratio of flicker amplitudes in the left and right eyes was varied across blocks and was set to be 0, 0.25, 0.5, 0.75 or 1 (9 distinct conditions). The standard and target discs were displayed for 1 second each, with an interstimulus interval of 0.5 seconds. After both discs had appeared on screen, the participants had to indicate which interval they perceived as having the more intense flicker. The intervals were randomly ordered, and all discs flickered at a frequency of 2 Hz (two cycles in sine phase).

@@

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software [@Kassner2014], which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit a single sine-wave at the fundamental frequency, rather than using Fourier analysis [e.g. @Spitschan2014]. The Fourier approach is more robust to noise at other frequencies (which can make the phase and amplitude of a fitted sine wave unstable) and has been used in some previous studies [see @Barrionuevo2014; @Barrionuevo2016]. Additionally, it makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis [e.g. @Figueira2022].

@@

save(file=paste0(datadir,'Averagedata.RData'),list=c('cleanmeansP','cleanmeansPCI','cleanmasksP','cleanmasksPCI','cleanmeansEav','cleanmeansEavCI','cleanmasksEav','cleanmasksEavCI','cleanmeansE2av','cleanmeansE2avCI','cleanmasksE2av','cleanmasksE2avCI','cleanmeansE','cleanmeansECI','cleanmasksECI','cleanmeansE2','cleanmeansE2CI','cleanmasksE2','cleanmasksE2CI','meanspectraP','meanspectraPCI','meanspectraE','meanspectraECI','meanwavesP','meanwavesE','meanwavesPCI','meanwavesECI','electrodes','allmeansP','allmeansE','allmeansE2','allmasksP','allmasksE','allmasksE2'))

@@

alldata <- NULL
allgroupA <- NULL
allgroupB <- NULL
allP <- NULL
counter <- 0
for (p in 1:30){
  for (lev in 1:5){
    counter <- counter + 1
    alldata[counter] <- allmasksP[p,4,lev]
    allgroupA[counter] <- lev
    allgroupB[counter] <- 1
    allP[counter] <- p

    counter <- counter + 1
    alldata[counter] <- allmasksP[p,5,lev]
    allgroupA[counter] <- lev
    allgroupB[counter] <- 2
    allP[counter] <- p
  }
}

monbinPstatsX2 <- anovacirc2way(alldata,allgroupA,allgroupB,allP)

alldata <- NULL
allgroupA <- NULL
allgroupB <- NULL
allP <- NULL
counter <- 0
for (p in 1:30){
  for (lev in 1:5){
    counter <- counter + 1
    alldata[counter] <- allmasksP[p,4,lev]
    allgroupA[counter] <- lev
    allgroupB[counter] <- 1
    allP[counter] <- p

mondichPstatsX2 <- anovacirc.test(alldata,allgroupA,allP)

@@

```{r pupildata, fig.cap="Summary of pupillometry results for N=30 participants. Panel (a) shows a group average waveform for binocular presentation (low pass filtered at 5Hz), with the driving signal plotted at the foot. Negative values indicate constriction relative to baseline, and positive values indicate dilation. Panel (b) shows the average Fourier spectrum (absolute amplitude values), with an inset image illustrating the stimulus appearance (upper right). Panels (c,d) show contrast response functions for pupil diameter at 2Hz for different conditions. Panel (e) shows contrast response functions at 1.6Hz for three conditions. Shaded regions and error bars indicate bootstrapped standard errors.", fig.align="center", echo=FALSE}

@@

Figure \@ref(fig:pupildata)e shows responses at 1.6Hz, for the same conditions, as well as for a condition in which a monocular stimulus flickered at 1.6Hz (grey circles). Here we find strong suppression in both the binocular cross (purple triangles) and dichoptic cross (yellow triangles) conditions. In the binocular cross condition, the amplitudes are reduced relative to the monocular condition (grey circles): F(`r monbinPstatsX2$dfAB`,`r monbinPstatsX2$dfR`) = `r round(monbinPstatsX2$FratioAB,digits=2)`, $p$ < 0.001). In the dichoptic cross condition, increasing the 2Hz target contrast suppresses the response to the 1.6Hz mask, and the function decreases [see e.g. @Busse2009]: F(`r mondichPstatsX2$dfM`,`r mondichPstatsX2$dfR`) = `r round(mondichPstatsX2$Fratio,digits=2)`, $p$ < 0.001).

@@

```{r BSratios, fig.cap="Ratio of binocular to monocular response for three data types. These were calculated by dividing the binocular response by the monocular response at each contrast level, using the data underlying Figures 1c, 2c and 2f. Each value is the average of ratio across N=30 participants, and error bars indicate bootstrapped standard errors.", fig.align="center", out.width = '50%', echo=FALSE}

@@

```{r TFdata, fig.cap="Binocular facilitation at different temporal frequencies, measured using EEG. Panel (a) shows Fourier spectra for responses to binocular flicker at 5 different frequencies (offset vertically for clarity). Panel (b) shows the response at each stimulation frequency for monocular (red circles) and binocular (blue squares) presentation. Panel (c) shows the ratio of binocular to monocular responses. Error bars and shaded regions indicate bootstrapped standard errors across N=12 participants.", fig.align="center", echo=FALSE}

@@

```{r matchingdata, fig.cap="Contrast matching functions. Dotted and dashed lines are predictions of canonical summation models involving linear combination (dotted) or a winner-take-all rule (dashed). Error bars indicate the standard error across participants (N=10), and are constrained along radial lines converging at the origin. Note that, for the {48\\%} match, the data point on the x axis falls higher than {100\\%} contrast. This is because the psychometric function fits for some individuals were interpolated such that the PSE fell above {100\\%}, shifting the mean slightly above that value.", out.width = '50%', fig.align="center", echo=FALSE}

@@

Pupil size affects the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2% of overall diameter [e.g. @Spitschan2014]. It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms [e.g. @Bradley2008]. We think it unlikely that these temporally coarse processes would have a differential effect on e.g. monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the pupil and perceptual pathways are effectively distinct, but hope to investigate this directly in future neuroimaging work.
